### 2023-06-22 introduction.tex
up

{\bf Context.} 
Resilience is a key notion for improving safety of unperfect systems and resilience engineering is a paradigm for safety management that focuses on systems coping with complexity and balancing productivity with safety (\cite{challenges}). Some systems are subjects at frequent intervals to accidents, attacks or changes. Think for instance of a supply chain, or an airport’s air trafic control. In such cases, a question that arises is that of whether the system can return to its normal (safe) behavior after an accident or attack
pushed it towards some kind of ‘error state’ and, if it can, whether it can perform the return in a satisfactory timeframe. 
%%%%%%%%
\alain{fait double emploi avec l'exemple si on le développe ce que je ne suis plus sûr...:  In the exemple of air trafic control, an accident i.e. for instance a delay in the arrival of a plane can lead to conflicts of resolution (i.e. landing) which lead to prompting planes to remain on standby. Since their fuel quantity is limited it is hence pretty important to be able to know that trafic will return to its normal pace quickly enough. We formalize this as the resilience problem. 
Resilience is then the capacity of a system that fall in a bad set of states to leave this set and to reach a safe set of states. \\

@@

{\bf Home-spaces}
 In 1986, Memmi and Vautherin introduced the notion of home-space \cite{DBLP:conf/ac/MemmiV86} for a system $\mathscr{S} = (S,\rightarrow )$ with an initial state $s_0$ : a subset $H \subseteq S$ is an \emph{home-space}  if 
%			from all reachable states from $s_0$, it is always possible to reach $H$ 
$\post^*(s_0) \subseteq \pred^*(H)$; it could be easily generalized for two subsets $X,H$, we say that $H$ is an home-space for $X$ if $\post^*(X) \subseteq \pred^*(H)$. In 1989, de Frutos Escrig and Johnen proved that the home-space problem (for $X$ a singleton and $H$ a finite union of linear sets with the same period) was decidable for VASS (only published as an internal report). In 2023, Jancar and Leroux proved the decidability of the (complete) semilinear home-space problem ($X$ and $H$ are both semilinear)  in VASS \cite{DBLP:journals/corr/abs-2207-02697}.
%%%%%

{\bf Resiliences.}  
%		We now give the definition of two kinds of resilience: the (general) resilience and the state-resilience. 
The (general) resilience property for a transition system $\mathscr{S} = (S,\rightarrow )$ and two subsets of states $Bad,Safe \subseteq S$ can be stated as an home-space problem : $\mathscr{S}$ is (Bad,Safe)-resilient if $\post^*(Bad) \subseteq \pred^*(Safe)$; 
similarly, $\mathscr{S}$ is (Bad,Safe)-state-resilient for an initial state $s_0$ if $\post^*(s_0) \cap Bad \subseteq \pred^*(Safe)$. Resilience implies state-resilience but a system can be state-resilient but not resilient. 
We will study three decidability questions for each of type of resilience.
The resilience problem is to decide whether a transition system $\mathscr{S} = (S,\rightarrow )$ is (Bad,Safe)-resilient (for two given subsets of states $Bad,Safe \subseteq S$).
The $k$-resilience problem (kRP) is to decide whether $\post^*(Bad) \subseteq \pred^{\leq k}(Safe)$ (for a given $k$) and 
the bounded resilience problem (BRP) is to decide whether there exists an $k$ such that $\mathscr{S}$ is (Bad,Safe)-$k$-resilient. We adapt these three definitions to state-resilience.

@@

Safety requires that only one aircraft at a time attempts to land. The role of the control tower is to ensure this;
there are two possible choices: aircraft $A$ waits for aircraft $B$ to land
or vice-versa. To make its choice known to the aircrafts, the control tower will use the counters. 
We assume that the VASS is
lossy, i.e. 
 the contents of a counter can spontaneously get lower at any time.
 Lossiness is represented by using transitions decreasing any counter by one from every state. 
% We ask however that the system is not too lossy. We add the condition that for every prefix of an execution, the number of ``lossiness transition'' for a counter is less than nine tenth of the number of transitions incrementing said counter.
In this example, the safe, desirable state would be one where the aircrafts have both landed, i.e.
$\Safe = \{(q_{A,1}, q_{M,i}, q_{B,1})(\textbf{v}) \mid \textbf{v} \in \N^4, i \in \N\}$, and $\Bad$ is the complement of $\Safe$. 
% We remark that $\Safe$ and $\Bad$ are here both upward and downward closed.
Resilience here expresses that the aircrafts will both land, while explicit resilience or $k$-resilience asks whether they will do so in at most $k$ steps. 
Explicit resilience is of interest since in general we want to ensure the safe space will
be reached in
the least amount of steps possible. 
In this exemple, for instance, aircrafts cannot simply wait in the air an infinite amount of time.
With one out of ten increment of a counter guarranteed to not be lost, $8$-resilience hold in the example.
\end{example}

>>>>

{\bf Context.} 
Resilience is a key notion for improving safety of unperfect systems and resilience engineering is a paradigm for safety management that focuses on systems coping with complexity and balancing productivity with safety~\cite{challenges}. Some systems are subjects at frequent intervals to accidents, attacks or changes. Think for instance of a supply chain, or an airport’s air trafic control. In such cases, a question that arises is that of whether the system can return to its normal (safe) behavior after an accident or attack
pushed it towards some kind of ‘error state’ and, if it can, whether it can perform the return in a satisfactory timeframe. 
%%%%%%%%
\alain{fait double emploi avec l'exemple si on le développe ce que je ne suis plus sûr...:  In the exemple of air trafic control, an accident i.e. for instance a delay in the arrival of a plane can lead to conflicts of resolution (i.e. landing) which lead to prompting planes to remain on standby. Since their fuel quantity is limited it is hence pretty important to be able to know that trafic will return to its normal pace quickly enough. We formalize this as the resilience problem. 
Resilience is then the capacity of a system that fall in a bad set of states to leave this set and to reach a safe set of states. \\

@@

{\bf Home-spaces}
 In 1986, Memmi and Vautherin introduced the notion of home-space~\cite{DBLP:conf/ac/MemmiV86} for a system $\mathscr{S} = (S,\rightarrow )$ with an initial state $s_0$ : a subset $H \subseteq S$ is an \emph{home-space}  if 
%			from all reachable states from $s_0$, it is always possible to reach $H$ 
$\post^*(s_0) \subseteq \pred^*(H)$; it could be easily generalized for two subsets $X,H$, we say that $H$ is an home-space for $X$ if $\post^*(X) \subseteq \pred^*(H)$. In 1989, de Frutos Escrig and Johnen proved that the home-space problem (for $X$ a singleton and $H$ a finite union of linear sets with the same period) was decidable for VASS (only published as an internal report). In 2023, Jancar and Leroux proved the decidability of the (complete) semilinear home-space problem ($X$ and $H$ are both semilinear)  in VASS \cite{DBLP:journals/corr/abs-2207-02697}.
%%%%%

{\bf Resiliences.}  
%		We now give the definition of two kinds of resilience: the (general) resilience and the state-resilience. 
The (general) resilience property for a transition system $\mathscr{S} = (S,\rightarrow )$ and two subsets of states $\Bad,\Safe \subseteq S$ can be stated as an home-space problem : $\mathscr{S}$ is ($\Bad,\Safe$)-resilient if $\post^*(\Bad) \subseteq \pred^*(\Safe)$; 
similarly, $\mathscr{S}$ is (\Bad,\Safe)-state-resilient for an initial state $s_0$ if $\post^*(s_0) \cap \Bad \subseteq \pred^*(\Safe)$. Resilience implies state-resilience but a system can be state-resilient but not resilient. 
We will study three decidability questions for each of type of resilience.
The resilience problem is to decide whether a transition system $\mathscr{S} = (S,\rightarrow )$ is ($\Bad,\Safe$)-resilient (for two given subsets of states $\Bad,\Safe \subseteq S$).
The $k$-resilience problem (kRP) is to decide whether $\post^*(\Bad) \subseteq \pred^{\leq k}(\Safe)$ (for a given $k$) and 
the bounded resilience problem (BRP) is to decide whether there exists an $k$ such that $\mathscr{S}$ is ($\Bad,\Safe$)-$k$-resilient. We adapt these three definitions to state-resilience.

@@

Safety requires that only one aircraft at a time attempts to land. The role of the control tower is to ensure this;
there are two possible choices: aircraft $A$ waits for aircraft $B$ to land
or vice-versa. To make its choice known to the aircrafts, the control tower will use the counters. 
We assume that the VASS is
lossy, i.e. 
 the contents of a counter can spontaneously get lower at any time.
 Lossiness is represented by using transitions decreasing any counter by one from every state. 
% We ask however that the system is not too lossy. We add the condition that for every prefix of an execution, the number of ``lossiness transition'' for a counter is less than nine tenth of the number of transitions incrementing said counter.
In this example, the safe, desirable state would be one where the aircrafts have both landed, i.e.
$\Safe = \{(q_{A,1}, q_{M,i}, q_{B,1})(\textbf{v}) \mid \textbf{v} \in \N^4, i \in \N\}$, and $\Bad$ is the complement of $\Safe$. 
% We remark that $\Safe$ and $\Bad$ are here both upward and downward closed.
Resilience here expresses that the aircrafts will both land, while explicit resilience or $k$-resilience asks whether they will do so in at most $k$ steps. 
Explicit resilience is of interest since in general we want to ensure the safe space will
be reached in
the least amount of steps possible. 
In this exemple, for instance, aircrafts cannot simply wait in the air an infinite amount of time. Fortunately, $8$-resilience hold in the example.
\end{example}

### 2023-06-22 resilience.tex
up

% Given a set of states $S$, with two subsets $\Bad$ and $\Safe$, 
It is useful to note that in a sequence of states% of $S$
, every prefix is always 
either in the case that the latest visited $\Bad$ state has already been followed by a 
$\Safe$ state
or in the case that it still has to. 
Thus, \emph{in context}, there is no true ``neutral'' state: either a neutral state is contextually ``$\Safe$'' because it follows a visit in a $\Safe$ state, either it is contextually ``$\Bad$'' because it follows a visit in a $\Bad$ state and has, accordingly, to reach a $\Safe$ one.
Hence we will consider in the following not only that $\Bad$ and $\Safe$ are disjoint but
% w.l.o.g.
	also
that $\Bad$ is the complement of $\Safe$,
in which case
if $\Bad$ is downward-closed then $\Safe$ is upward-closed and vice-versa.
Cases where $\Bad$ or $\Safe$ are not complements of each other, and share closure properties, i.e. are both downward-closed or both upward-closed,
have also been studied, 
see 
Appendix~\ref{case down down}
and
Appendix~\ref{case up up} for results.

>>>>

It is useful to note that in a sequence of states% of $S$
, every prefix is always 
either in the case that the latest visited $\Bad$ state has already been followed by a 
$\Safe$ state \-- and thus the latest state is contextually ``$\Safe$'' \--
or in the case that it still has to \-- and thus the latest state is contextually ``$\Bad$''. 
Thus, \emph{in context}, there is no true ``neutral'' state.
Hence we will consider in the following not only that $\Bad$ and $\Safe$ are disjoint but
% w.l.o.g.
	also
that $\Bad$ is the complement of $\Safe$,
in which case
if $\Bad$ is downward-closed then $\Safe$ is upward-closed and vice-versa.
Cases where $\Bad$ or $\Safe$ are not complements of each other, and share closure properties, i.e. are both downward-closed or both upward-closed,
have also been studied, 
see 
Appendix~\ref{case down down}
and
Appendix~\ref{case up up} for results.
